---
output: github_document
always_allow_html: yes
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# faux

It is useful to be able to simulate data with a specified structure. The `faux` package provides some functions to make this process easier.

## Installation

You can install the released version of faux from
[GitHub](https://github.com/debruine/faux) with:

``` r
devtools::install_github("debruine/faux")
```

## Examples

```{r libraries, message=FALSE}
library(tidyverse)
library(faux)
```


### rnorm_multi

This function makes multiple normally distributed vectors with specified parameters and relationships. 

For example, the following creates a sample that has 100 observations of 3 variables, drawn from a population where where A correlates with B and C with r = 0.5, and B and C correlate with r = 0.25. A has a mean of 0 and SD of 1, while B and C have means of 20 and SDs of 5.

```{r set-seed, include=FALSE}
set.seed(200)
```


```{r multirnorm-example}

dat <- rnorm_multi(n = 100, 
                  cors = c(0.5, 0.5, 0.25), 
                  mu = c(0, 20, 20),
                  sd = c(1, 5, 5),
                  varnames = c("A", "B", "C"),
                  empirical = FALSE)

```


`r check_sim_stats(dat, usekable = T)`
Table: Sample stats

#### Specify `cors`

You can specify the correlations in one of four ways:

* A single r for all pairs
* A vars by vars matrix
* A vars\*vars length vector
* A vars\*(vars-1)/2 length vector

##### One Number

If you want all the pairs to have the same correlation, just specify a single number. 

```{r}
bvn <- rnorm_multi(100, 5, .3, varnames = letters[1:5])
```

`r check_sim_stats(bvn, usekable = T)`
Table: Sample stats from a single rho

##### Matrix

If you already have a correlation matrix, such as the output of `cor()`, you can specify the simulated data with that.

```{r vvmatrix}
cmat <- cor(iris[,1:4])
bvn <- rnorm_multi(100, 4, cmat, 
                  varnames = colnames(cmat))
```

`r check_sim_stats(bvn, usekable = T)`
Table: Sample stats from a correlation matrix

##### Vector (vars\*vars)

You can specify your correlation matrix by hand as a vars\*vars length vector, which will include the correlations of 1 down the diagonal.

```{r}
cmat <- c(1, .3, .5,
          .3, 1, 0,
          .5, 0, 1)
bvn <- rnorm_multi(100, 3, cmat, 
                  varnames = c("first", "second", "third"))
```

`r check_sim_stats(bvn, usekable = T)`
Table: Sample stats from a vars\*vars vector

##### Vector (vars\*(vars-1)/2)

You can specify your correlation matrix by hand as a vars\*(vars-1)/2 length vector, skipping the diagonal and lower left duplicate values.

```{r}
rho1_2 <- .3
rho1_3 <- .5
rho1_4 <- .5
rho2_3 <- .2
rho2_4 <- 0
rho3_4 <- -.3
cmat <- c(rho1_2, rho1_3, rho1_4, rho2_3, rho2_4, rho3_4)
bvn <- rnorm_multi(100, 4, cmat, 
                  varnames = letters[1:4])
```

`r check_sim_stats(bvn, usekable = T)`
Table: Sample stats from a (vars\*(vars-1)/2) vector

#### empirical

If you want your samples to have the *exact* correlations, means, and SDs you entered, set `empirical` to TRUE.

```{r}
bvn <- rnorm_multi(100, 5, .3, 
                  varnames = letters[1:5], 
                  empirical = T)
```

`r check_sim_stats(bvn, usekable = T)`
Table: Sample stats with empirical = TRUE

### simdf

This function produces a dataframe with the same distributions and correlations as an existing dataframe. It only returns numeric columns and simulates all numeric variables from a continuous normal distribution (for now).

For example, here is the relationship between speed and distance in the built-in dataset `cars`. 

```{r plot-cars-orig, fig.cap="Original cars dataset"}
cars %>%
  ggplot(aes(speed, dist)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

You can create a new sample with the same parameters and 500 rows with the code `simdf(cars, 500)`.

```{r plot-cars-sim, fig.cap="Simulated cars dataset"}
simdf(cars, 500) %>%
  ggplot(aes(speed, dist)) + 
    geom_point() +
    geom_smooth(method = "lm")
```

#### Grouping Variables

You can also optionally add grouping variables. For example, here is the relationship between sepal length and width in the built-in dataset `iris`.

```{r plot-iris-orig, fig.cap="Original iris dataset"}
iris %>%
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

And here is a new sample with 50 observations of each species, made with the code `simdf(iris, 100, "Species")`.

```{r plot-iris-sim, fig.cap="Simulated iris dataset"}
simdf(iris, 50, "Species") %>%
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

For now, the function only creates new variables sampled from a continuous normal distribution. I hope to add in other sampling distributions in the future. So you'd need to do any rounding or truncating yourself.

```{r plot-iris-sim-round, fig.cap="Simulated iris dataset (rounded)"}
simdf(iris, 50, "Species") %>%
  mutate_if(is.numeric, round, 1) %>%
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

### pos_def_limits

Not all correlation matrices are possible. For example, if variables A and B are correlated with r = 1.0, then the correlation between A and C can only be exactly equal to the correlation between B and C.

The function `pos_def_limits()` lets you know what the range of values is for a correlation matrix with one missing value. The correlation values are entered just from the top right triangle of the matrix, with a single `NA` for the missing value.

```{r pos_def_limits1}
lims <- pos_def_limits(.8, .2, NA)
```

`r knitr::kable(lims)`

For example, if r~AB~ = 0.8 and r~AC~ = 0.2, then `r lims$min` <= r~BC~ <= `r lims$max`.

If you enter a correlation matrix that contains impossible combinations, your limits will be `NA`.

```{r pos_def_limits-na}
lims <- pos_def_limits(.8, .2,  0,
                          -.5, NA,
                               .2)
```

`r knitr::kable(lims)`

If you have a full matrix and want to know if it positive definite, you can use the following code:

```{r is_pos_def}
c(.2, .3, .4, .2,
      .3, -.1, .2,
           .4, .5,
               .3) %>%
  cormat_from_triangle() %>%
  is_pos_def()
```

```{r not_pos_def}
matrix(c(1, .3, -.9, .2,
        .3,  1,  .4, .5,
       -.9, .4,   1, .3,
        .2, .5,  .3,  1), 4) %>%
  is_pos_def()
```

### check_sim_stats

If yu want to check your simulated stats or just describe an existing dataset, use `check_sim_stats()`.

```{r}
check_sim_stats(iris)
```

You can also group your data and change the digits to round. Display the table using `knitr::kable()` by setting `usekable` to `TRUE` (remember to set `results='asis'` in the chunk header.

```{r, results='asis'}
check_sim_stats(iris, 
                grp_by = "Species", 
                digits = 3, 
                usekable = TRUE)
```
