---
title: "Contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# run a simple analysis 
ctable <- function(df, predictor = "pet") {
  a <- paste("y ~", predictor) %>%
    formula() %>%
    lm(df)
  
  paste(
    "<table><thead><tr><th>Contrasts</th><th>lm(y ~ pet, df)</th></thead>
  <tbody><tr><td>\n",
    knitr::kable(contrasts(df[[predictor]]), format = "html"),
    "\n</td><td>\n",
    knitr::kable(broom::tidy(a), format = "html"),
    "\n</td></tr></tbody></table>\n"
  )
}

options(digits = 3)
set.seed(8675309)

library(faux)
```


Contrasts beyond treatment coding and 2-level deviation coding have always confused me. I try to design all my own studies to never have more than two levels per categorical factor, but students in my data skills class are always trying to analyse data with three-level categories, or worse. 

I thought it was just me, but some recent Twitter discussion showed me I'm not alone. A bit of research showed me that there's a serious jingle-jangle problem with contrasts. Here are some of the explainers that I found useful (thanks to everyone who recommended them). 

* [Contrasts in R](https://marissabarlaz.github.io/portfolio/contrastcoding/) by Marissa Barlaz
* [Coding categorical predictor variables in factorial designs](https://talklab.psy.gla.ac.uk/tvw/catpred/) by Dale Barr
* [Contrast Coding in Multiple Regression Analysis](https://doi.org/10.6339/JDS.2010.08(1).563) by Matthew Davis
* [R Library Contrast Coding Systems for categorical variables](https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/) by UCLA Statistical Consulting
* [Rosetta store: Contrasts](https://gamlj.github.io/rosetta_contrasts.html) by GAMLj


## Terminology

Contrasts often have multiple names. The names I'm using try to maintain the relationship with the base R function, although I'm not sure what I should call deviation coding, as it has so many conflicting names.

| My name | Other names | R function | faux function |
|:--------|:------------|:-----------|:--------------|
| [Treatment](#treatment) | [Treatment](https://talklab.psy.gla.ac.uk/tvw/catpred/#what-are-the-key-coding-schemes), [Dummy](https://marissabarlaz.github.io/portfolio/contrastcoding/#dummy-coding), [Dummy](https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/#dummy), [Simple](https://gamlj.github.io/rosetta_contrasts.html#Contrast:_Simple__Dummy) | `contr.treatment` | `contr_code_treatment` |
| [Deviation](#deviation) | [Deviation](https://talklab.psy.gla.ac.uk/tvw/catpred/#what-are-the-key-coding-schemes), [Simple](https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/#SIMPLE), [Contrast](https://marissabarlaz.github.io/portfolio/contrastcoding/#contrast-coding) | `contr.treatment - 1/k` | `contr_code_deviation` |
| [Difference](#difference) | [Forward/Backward](https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/#forward), [Contrast](https://doi.org/10.6339/JDS.2010.08(1).563), [Repeated](https://gamlj.github.io/rosetta_contrasts.html#Contrast:_Repeated) | `MASS::contr.sdif` | `contr_code_difference` |
| [Sum](#sum) | [Sum](https://talklab.psy.gla.ac.uk/tvw/catpred/#what-are-the-key-coding-schemes), [Sum](https://marissabarlaz.github.io/portfolio/contrastcoding/#sum-coding), [Deviation](https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/#SIMPLE), [Deviation](https://gamlj.github.io/rosetta_contrasts.html#Contrast:_Deviation), [Effects](https://doi.org/10.6339/JDS.2010.08(1).563) | `contr.sum` | `contr_code_sum` |
| [Helmert](#helmert) | [Reverse Helmert](https://marissabarlaz.github.io/portfolio/contrastcoding/#reverse-helmert-coding), [Reverse Helmert](https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/#reverse), [Difference](https://gamlj.github.io/rosetta_contrasts.html#Contrast:_Difference) | `contr.helmert / column_i` | `contr_code_helmert` |
| [Polynomial](#poly) | [Polynomial](https://gamlj.github.io/rosetta_contrasts.html#Contrast:_Polynomial), [Orthogonal](https://stats.idre.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/#ORTHOGONAL), [Trend](https://doi.org/10.6339/JDS.2010.08(1).563) | `contr.poly` | `contr_code_poly` |


## Faux Function Examples

**These functions are under development**

First, we'll set up a simple experimental design and analyse it with `lm()`. I set `empirical = TRUE` to make interpreting the estimates easier. 


```{r}
df <- sim_design(between = list(pet = c("cat", "dog", "ferret")), 
                 n = c(50, 100, 200), mu = c(2, 4, 9), empirical = TRUE)
```


Notice that the default contrast is treatment coding, with cat as the baseline condition. The estimate for the intercept is the mean value for cats ($2$), while the term `petdog` is the mean value for dogs minus cats ($4 - 2$), and the term `petferret` is the mean value for ferrets minus cats ($9 - 2$).

`r ctable(df)`

In all the subsequent examples, try putting into words what the estimates for the intercept and terms mean in relation to the mean values for each group in the data.

### Treatment Coding {#treatment}

Treatment coding (also called dummy coding) sets the mean of the reference group as the intercept. This is the same type of coding as the default for factors shown above (unless you change your default using `options()`), but with clearer term labels. 

```{r}
df$pet <- contr_code_treatment(df$pet)
```

`r ctable(df)`

Each contrast compares one level with the reference level, which defaults to the first level, but you can set with the `base` argument. Now the intercept estimates the mean value for dogs ($4$).

```{r}
df$pet <- contr_code_treatment(df$pet, base = "dog")
```

`r ctable(df)`

Set the reference level to the third level ("ferret").

```{r}
df$pet <- contr_code_treatment(df$pet, base = 3)
```

`r ctable(df)`


### Deviation Coding {#deviation}

Deviation coding is similar to treatment coding, but sets the grand mean as the intercept. Each contrast compares one level with a reference level.

Deviation coding with the first level ("cat") as the default base. Now the intercept is the grand mean, which is the mean of the three group means ($(2 + 4 + 9)/3$). Notice that this is different from the mean value of y in our dataset (`r mean(df$y)`, since the number of pets in each group is unbalanced. The term `pet_dog-cat` is the mean value for dogs minus cats ($4 - 2$) and the term `pet_ferret-cat` is the mean value for ferrets minus cats ($9 - 2$).

```{r}
df$pet <- contr_code_deviation(df$pet)
```

`r ctable(df)`

Deviation coding with "dog" as the base. How does the interpretation of the terms change?

```{r}
df$pet <- contr_code_deviation(df$pet, base = "dog")
```

`r ctable(df)`

Deviation coding with the third level ("ferret") as the base.

```{r}
df$pet <- contr_code_deviation(df$pet, base = 3)
```

`r ctable(df)`

### Difference Coding {#difference}

A slightly different form of contrast coding is difference coding, also called forward, backward, or successive differences coding. It compares each level to the previous one, rather than to a baseline level.

```{r}
df$pet <- contr_code_difference(df$pet)
```

`r ctable(df)`

If you want to change which levels are compared, you can re-order the factor levels.

```{r}
df$pet <- contr_code_difference(df$pet, levels = c("ferret", "cat", "dog"))
```

`r ctable(df)`


```{r, include = FALSE}
# get levels back to normal
df$pet <- contr_code_difference(df$pet, levels = c("cat", "dog", "ferret"))
```

### Sum Coding {#sum}

Sum coding also sets the grand mean as the intercept. Each contrast compares one level with the grand mean. Therefore, the estimate for `pet_cat-intercept` is the difference between the mean value for cats and the grand mean ($2 - 5$).

```{r}
df$pet <- contr_code_sum(df$pet)
```

`r ctable(df)`

You can't compare all levels with the grand mean, and have to omit one level. This is the last level by default, but you can change it with the `omit` argument.

```{r}
df$pet <- contr_code_sum(df$pet, omit = "dog")
```

`r ctable(df)`

Omit the first level ("cat").

```{r}
df$pet <- contr_code_sum(df$pet, omit = 1)
```

`r ctable(df)`


### Helmert Coding {#helmert}

Helmert coding sets the grand mean as the intercept. Each contrast compares one level with the mean of previous levels. This coding is somewhat different than the results from `stats::contr.helmert()` to make it easier to interpret the estimates. Thus, `pet_ferret-cat.dog` is the mean value for ferrets minus the mean value for cats and dogs averaged together ($9-(2+4)/2$).

```{r}
df$pet <- contr_code_helmert(df$pet)
```

`r ctable(df)`

You can change the comparisons by reordering the levels.

```{r}
df$pet <- contr_code_helmert(df$pet, levels = c("ferret", "dog", "cat"))
```

`r ctable(df)`

### Polynomial Coding {#poly}

Polynomial coding is the default for ordered factors in R. We'll set up a new data simulation with five ordered times.

```{r}
df <- sim_design(list(time = 1:5),
                 mu = 1:5 * 0.25 + (1:5 - 3)^2 * 0.5,
                 sd = 5, long = TRUE)
```

The function `contr_code_poly()` uses `contr.poly()` to set up the polynomial contrasts for the linear (`^1`), quadratic (`^2`), cubic (`^3`), and quartic (`^4`) components.

```{r}
df$time <- contr_code_poly(df$time)
```

`r ctable(df, "time")`

## Add Contrasts

The function `add_contrasts()` lets you add contrasts to a column in a data frame and also adds new columns for each contrast (unless `add_cols = FALSE`). This is especially helpful if you want to test only a subset of the contrasts.

```{r}
df <- sim_design(list(time = 1:5),
                 mu = 1:5 * 0.25 + (1:5 - 3)^2 * 0.5,
                 sd = 5, long = TRUE, plot = FALSE) %>%
  add_contrast("time", "poly")
```


```{r, results='hide'}
# test only the linear and quadratic contrasts
lm(y ~ `time^1` + `time^2`, df) %>% broom::tidy()
```

```{r, echo = FALSE}
lm(y ~ `time^1` + `time^2`, df) %>% 
  broom::tidy() %>%
  knitr::kable()
```

You can set `colnames` to change the default column names for the contrasts. This can be useful if you want to add different codings for the same factor or if the default names are too long.

```{r}
btwn <- list(condition = c("control", "experimental")) 

df <- sim_design(between = btwn, n = 1, plot = FALSE) %>%
  add_contrast("condition", "treatment", colnames = "cond.tr") %>%
  add_contrast("condition", "deviation", colnames = "cond.dev") %>%
  add_contrast("condition", "difference", colnames = "cond.dif") %>%
  add_contrast("condition", "sum", colnames = "cond.sum") %>%
  add_contrast("condition", "helmert", colnames = "cond.hmt") %>%
  add_contrast("condition", "poly", colnames = "cond.poly")
```

`r knitr::kable(df)`

However, if a new column has a duplicate name to an existing column, `add_contrast()` will automatically add a contrast suffix to the new column. 

```{r}
btwn <- list(pet = c("cat", "dog", "ferret")) 

df <- sim_design(between = btwn, n = 1, plot = FALSE) %>%
  add_contrast("pet", "treatment") %>% 
  add_contrast("pet", "deviation") %>%
  add_contrast("pet", "difference") %>%
  add_contrast("pet", "sum") %>%
  add_contrast("pet", "helmert") %>%
  add_contrast("pet", "poly")
```

`r knitr::kable(df, digits = 2, )`
