---
title: "Introduction to multisim"
author: "Lisa DeBruine"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to multisim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(dplyr)
library(tidyr)
library(purrr)
library(multisim)
```


It is useful to be able to simulate data with a specified structure. The `multisim` package provides some functions to make this process easier.


## multirnorm

This function makes multiple normally distributed vectors with specified parameters and relationships. 

For example, the following creates a sample that has 100 observations of 3 variables, drawn from a population where where A correlates with B and C with r = 0.5, and B and C correlate with r = 0.25. A has a mean of 0 and SD of 1, while B and C have means of 20 and SDs of 5.

```{r set-seed, include=FALSE}
set.seed(200)
```


```{r multirnorm-example}

dat <- multirnorm(n = 100, 
                  cors = c(0.5, 0.5, 0.25), 
                  mu = c(0, 20, 20),
                  sd = c(1, 5, 5),
                  varnames = c("A", "B", "C"),
                  empirical = FALSE)

```


`r checkstats(dat, usekable = T)`
Table: Sample stats

### Specify `cors`

You can specify the correlations in one of four ways:

* A single r for all pairs
* A vars by vars matrix
* A vars\*vars length vector
* A vars\*(vars-1)/2 length vector

#### One Number

If you want all the pairs to have the same correlation, just specify a single number. 

```{r}
bvn <- multirnorm(100, 5, .3, varnames = letters[1:5])
```

`r checkstats(bvn, usekable = T)`
Table: Sample stats from a single rho

#### Matrix

If you already have a correlation matrix, such as the output of `cor()`, you can specify the simulated data with that.

```{r vvmatrix}
cmat <- cor(iris[,1:4])
bvn <- multirnorm(100, 4, cmat, 
                  varnames = colnames(cmat))
```

`r checkstats(bvn, usekable = T)`
Table: Sample stats from a correlation matrix

#### Vector (vars\*vars)

You can specify your correlation matrix by hand as a vars\*vars length vector, which will include the correlations of 1 down the diagonal.

```{r}
cmat <- c(1, .3, .5,
          .3, 1, 0,
          .5, 0, 1)
bvn <- multirnorm(100, 3, cmat, 
                  varnames = c("first", "second", "third"))
```

`r checkstats(bvn, usekable = T)`
Table: Sample stats from a vars\*vars vector

#### Vector (vars\*(vars-1)/2)

You can specify your correlation matrix by hand as a vars\*(vars-1)/2 length vector, skipping the diagonal and lower left duplicate values.

```{r}
rho1_2 <- .3
rho1_3 <- .5
rho1_4 <- .5
rho2_3 <- .2
rho2_4 <- 0
rho3_4 <- -.3
cmat <- c(rho1_2, rho1_3, rho1_4, rho2_3, rho2_4, rho3_4)
bvn <- multirnorm(100, 4, cmat, 
                  varnames = letters[1:4])
```

`r checkstats(bvn, usekable = T)`
Table: Sample stats from a (vars\*(vars-1)/2) vector

### empirical

If you want your samples to have the *exact* correlations, means, and SDs you entered, set `empirical` to TRUE.

```{r}
bvn <- multirnorm(100, 5, .3, 
                  varnames = letters[1:5], 
                  empirical = T)
```

`r checkstats(bvn, usekable = T)`
Table: Sample stats with empirical = TRUE
