---
title: "Simulate from Existing Data"
author: "Lisa DeBruine"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate from Existing Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
ggplot2::theme_set(ggplot2::theme_bw())
set.seed(8675309)
```

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(faux)
```

The `sim_df()` function produces a dataframe with the same distributions and correlations as an existing dataframe. It only returns numeric columns and simulates all numeric variables from a continuous normal distribution (for now).

For example, here is the relationship between speed and distance in the built-in dataset `cars`. 

```{r plot-cars-orig, fig.cap="Original cars dataset"}
cars %>%
  ggplot(aes(speed, dist)) + 
  geom_point() +
  geom_smooth(method = "lm", formula = "y~x")
```

You can create a new sample with the same parameters and 500 rows with the code `sim_df(cars, 500)`.

```{r plot-cars-sim, fig.cap="Simulated cars dataset"}
sim_df(cars, 500) %>%
  ggplot(aes(speed, dist)) + 
    geom_point() +
    geom_smooth(method = "lm", formula = "y~x")
```

## Between-subject variables

You can also optionally add between-subject variables. For example, here is the relationship between horsepower (`hp`) and weight (`wt`) for automatic (`am = 0`) versus manual (`am = 1`) transmission in the built-in dataset `mtcars`.

```{r plot-mtcars-orig, fig.cap="Original mtcars dataset"}
mtcars %>%
  mutate(transmission = factor(am, labels = c("automatic", "manual"))) %>%
  ggplot(aes(hp, wt, color = transmission)) +
  geom_point() +
  geom_smooth(method = "lm", formula = "y~x")
```

And here is a new sample with 50 observations of each.

```{r plot-iris-sim, fig.cap="Simulated iris dataset"}
sim_df(mtcars, 50 , between = "am") %>%
  mutate(transmission = factor(am, labels = c("automatic", "manual"))) %>%
  ggplot(aes(hp, wt, color = transmission)) +
  geom_point() +
  geom_smooth(method = "lm", formula = "y~x")
```

## Empirical

Set `empirical = TRUE` to return a data frame with *exactly* the same means, SDs, and correlations as the original dataset.

```{r}
exact_mtcars <- sim_df(mtcars, 50, between = "am", empirical = TRUE)
```


## Rounding

For now, the function only creates new variables sampled from a continuous normal distribution. I hope to add in other sampling distributions in the future. So you'd need to do any rounding or truncating yourself.

```{r plot-iris-sim-round, fig.cap="Simulated iris dataset (rounded)"}
sim_df(mtcars, 50, between = "am") %>%
  mutate(hp = round(hp),
         transmission = factor(am, labels = c("automatic", "manual"))) %>%
  ggplot(aes(hp, wt, color = transmission)) +
  geom_point() +
  geom_smooth(method = "lm", formula = "y~x")
```

