---
title: "Simulate by Design"
author: "Lisa DeBruine"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate by Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)

library(ggplot2)
library(dplyr)
library(tidyr)
library(faux)
theme_set(theme_bw())
set.seed(200)
```

The `sim_design()` function creates a dataset with a specific between- and within-subjects design. 


## Quick example

For example, the following creates a 2w*2b design with 100 observations in each cell. The between-subject factor is `pet` with twolevels of `cat` and `dog`. The within-subject factor is `time` with two levels of `day` and `night`. The mean for the `cat_day` cell is 10, the mean for the `cat_night` cell is 20, the mean for the `dog_day` cell is 15, and the mean for the `dog_night` cell is 25. All cells have a SD of 5 and all within-subject cells are correlated <code>r = 0.5</code>. The resulting data has exactly these values (set `empirical = FALSE` to sample from a population with these values).

```{r}
between <- list("pet" = c("cat", "dog"))
within <- list("time" = c("day", "night"))
mu <- data.frame(
  cat = c(10, 20),
  dog = c(15, 25),
  row.names = within$time
)
df <- sim_design(within, between, n = 100, mu = mu, sd = 5, r = 0.5, empirical = TRUE)
```

```{r echo = FALSE, result='asis'}
check_sim_stats(df, between = "pet", usekable = TRUE)
```

```{r, echo = FALSE, fig.width = 8, fig.height = 4}
gather(df, time, val, day:night) %>%
  ggplot(aes(pet, val, fill = pet)) +
  geom_violin(show.legend = FALSE) +
  geom_boxplot(width = .2, fill = "white", show.legend = FALSE) +
  facet_grid(~time) +
  scale_fill_manual(values = c("dodgerblue", "red"))
```

## Specification of design parameters

### Factor and level names

If you don't feel like naming your factors and levels, you can just put in a vector of levels. So you can make a quick 2w\*3w\*2b with the following code.

```{r}
df <- sim_design(within = c(2,3), between = c(2), n = 5, mu = 0, sd = 1, r = 0.5)
```

`r knitr::kable(df)`

You can specify between-subject and within-subject factors as a list of vectors where the item names are the factor labels and the vectors are the level labels.

```{r}
between <- list(
  "pet" = c("cat", "dog")
)
within <- list(
  "time" = c("day", "night")
)
```

You can also specify factors as a list of named vectors where the item names are the factor labels, the vector names are the level labels, and the vector items are the long labels you would use in a codebook or plot.

```{r}
between <- list(
  pet = c(cat = "Is a cat person", dog = "Is a dog person")
)
within <- list(
  time = c(day = "Tested during the day", night = "Tested at night")
)
```

The long label for factor levels isn't used yet, but will be used in future functions that create a codebook or plots.


### N, mean, SD

You can specify the Ns, means and standard deviations for each cell as a single number, list of vectors, or data frame. 

You usually want to specify `n` as a single number. This is N per cell, not total sample size.

```{r}
n <- 100 # n per cell, not total
```

The levels of the between-subject factors are the list names and the levels of the within-subject factors are the vector names. 

```{r}
mu <- list(
  cat = c(day = 10, night = 20),
  dog = c(day = 10, night = 30)
)
```

Alternatively, you can specify them as dataframes.

```{r}
sd <- data.frame(
  cat = c(3, 4),
  dog = c(3, 5),
  row.names = c("day", "night")
)
```

```{r, echo = FALSE, results='asis'}
knitr::kable(sd)
```


### Correlations

If you have any within-subject factors, you need to set the correlation for each between-cell. Here, we only have two levels of one within-subject factor, so can only set one correlation per between-cell.

```{r}
r <- list(
  cat = .5,
  dog = .6
)
```

If you set `empirical = TRUE`, you will get the *exact* means, SDs and correlations you specified. If you set `empirical = FALSE` or omit that argument, your data will be sampled from a population with those parameters, but your dataset will not have exactly those values (just on average).

```{r}
df <- sim_design(within, between, n = 100, 
                 mu = mu, sd = sd, r = r,
                 empirical = TRUE)
```

`r check_sim_stats(df, between = "pet", usekable = TRUE)`

## More factors

Here is a 2w\*3w\*2b\*2b example. When you have multiple within or between factors, you need to specify parameters by cell. Cell names are the level names, in the order they are listed in the `within` or `between` arguments, separated by underscores. 

Foe example, if you have one within-subject factor of condition with levels `con` and `inc`, and another within-subject factor of version wiith levels `easy`, `med`, and `hard`, your cell lables will be: `con_easy`, `inc_easy`, `con_med`, `inc_med`, `con_hard`, and `inc_hard`.

If you have any characters in your level names except letters and numbers, they will be replaced by a full stop (e.g., `my super-good level_name` will become `my.super.good.level.name`).

```{r}
within <- list(
  condition = c(con = "Mean of congruent trials", inc = "Mean of incongruent trials"),
  version = c(easy = "Easy", med = "Medium", hard = "Difficult")
)

between <- list(
  experience = c(novice = "Novice", expert = "Expert"),
  time = c(day = "Before 5pm", night = "After 5pm")
)

mu <- data.frame(
  row.names = c("con_easy", "inc_easy", "con_med", "inc_med", "con_hard", "inc_hard"),
  novice_day = c(10, 15, 12, 17, 14, 19),
  novice_night = c(11, 16, 13, 18, 15, 20),
  expert_day = c(9, 14, 11, 16, 13, 18),
  expert_night = c(10, 15, 12, 17, 14, 19)
)
```

You can set the correlation for each between-cell to a single number.

```{r}
r <- list(
  novice_day = 0.3,
  novice_night = 0.2,
  expert_day = 0.5,
  expert_night = 0.4
)
```

Or you can set the full correlation matrix with a vector or matrix. Since we have 6 within-cells, this is a 6x6 matrix or a vector of the upper right 15 values.

```{r}
# upper right triangle correlation specification
# inc and con have r = 0.5 within each difficultly level, 0.2 otherwise
#          ce,  ie,  cm,  im,  ch,  ih
triangle <-  c(0.5, 0.2, 0.2, 0.2, 0.2, #con_easy
                    0.2, 0.2, 0.2, 0.2, #inc_easy
                         0.5, 0.2, 0.2, #con_med
                              0.2, 0.2, #inc_med
                                   0.5) #con_hard
                                        #inc_hard

r <- list(
  novice_day = triangle,
  novice_night = triangle,
  expert_day = triangle,
  expert_night = triangle
)
```

You can set `frame_long = TRUE` to return the data frame in long format, which is usually easier for plotting. 

```{r}
df <- sim_design(within, between, n = 100, 
                 mu = mu, sd = 2, r = r, 
                 frame_long = TRUE)
```


```{r, echo = FALSE, fig.width = 8, fig.height = 5}
df %>%
  mutate(version = forcats::fct_relevel(version, names(within$version))) %>%
  mutate(condition = forcats::fct_relevel(condition, names(within$condition))) %>%
  mutate(experience = forcats::fct_relevel(experience, names(between$experience))) %>%
  mutate(time = forcats::fct_relevel(time, names(between$time))) %>%
  ggplot(aes(version, val, fill = condition, grp = condition)) +
  geom_violin(show.legend = FALSE) +
  geom_boxplot(width = .2, fill = "white", 
               position = position_dodge(width = 0.9), 
               show.legend = FALSE) +
  facet_grid(time~experience) +
  scale_fill_manual(values = c("dodgerblue", "red"))
```
