---
title: "Simulate from Existing Data"
author: "Lisa DeBruine"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulate from Existing Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 5,
  collapse = TRUE,
  comment = "#>"
)
library(ggplot2)
library(dplyr)
library(tidyr)
library(faux)
theme_set(theme_bw())
set.seed(200)
```

The `sim_df()` function produces a dataframe with the same distributions and correlations as an existing dataframe. It only returns numeric columns and simulates all numeric variables from a continuous normal distribution (for now).

For example, here is the relationship between speed and distance in the built-in dataset `cars`. 

```{r plot-cars-orig, fig.cap="Original cars dataset"}
cars %>%
  ggplot(aes(speed, dist)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

You can create a new sample with the same parameters and 500 rows with the code `sim_df(cars, 500)`.

```{r plot-cars-sim, fig.cap="Simulated cars dataset"}
sim_df(cars, 500) %>%
  ggplot(aes(speed, dist)) + 
    geom_point() +
    geom_smooth(method = "lm")
```

#### Grouping Variables

You can also optionally add grouping variables. For example, here is the relationship between sepal length and width in the built-in dataset `iris`.

```{r plot-iris-orig, fig.cap="Original iris dataset"}
iris %>%
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

And here is a new sample with 50 observations of each species, made with the code `sim_df(iris, 100, "Species")`.

```{r plot-iris-sim, fig.cap="Simulated iris dataset"}
sim_df(iris, 50, "Species") %>%
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

For now, the function only creates new variables sampled from a continuous normal distribution. I hope to add in other sampling distributions in the future. So you'd need to do any rounding or truncating yourself.

```{r plot-iris-sim-round, fig.cap="Simulated iris dataset (rounded)"}
sim_df(iris, 50, "Species") %>%
  mutate_if(is.numeric, round, 1) %>%
  ggplot(aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```
