---
title: "R to Tau"
author: "Lisa DeBruine"
date: "31/08/2020"
output: html_document
---
  
```{r}
knitr::opts_chunk$set(echo = TRUE, out.width = "100%", plot.width = 8, plot.height = 6)
library(lme4)
library(faux)
library(tidyverse)
```

## Function

Simulates data for DV `y` with `n_subj` subjects doing `n_trials` repeated trials across 2 conditions (`B`) with effect size `beta`. The subject random intercept SD is `tau_0` and the error SD is `sigma`.

Runs a mixed model with formula `y ~ B + (1 | id)` and a Cronbach's alpha.

```{r}
x <- function(n_subj = 10, 
              n_trials = 10, 
              beta = 0, 
              tau_0 = 0, 
              sigma = 1) {
  subj <- data.frame(
    id = 1:n_subj,
    T_0s = rnorm(n_subj, 0, tau_0),
    B = rep(c(-0.5, 0.5), n_subj/2)
  )
  
  dat <- crossing(subj, trial = 1:n_trials) %>%
    mutate(err = rnorm(nrow(.), 0, sigma),
           y = B*beta + T_0s + err)

  mod <- lmer(y ~ B + (1 | id), dat)
  
  # so much output to trash!
  trash <- suppressWarnings(
    suppressMessages(
      capture.output(
        alpha <- dat %>%
          select(id, trial, y) %>%
          spread(trial, y) %>%
          select(-id) %>%
          psych::alpha(warnings = FALSE, check.keys = FALSE) %>% summary()
  )))
  
  list(
    # params
    n_trials = n_trials,
    n_subj = n_subj,
    beta = beta,
    tau_0 = tau_0,
    sigma = sigma, 
    # psych::alpha output
    std.alpha = alpha$std.alpha,
    raw.alpha = alpha$raw_alpha,
    average_r = alpha$average_r,
    # lmer output
    B = fixef(mod)[[2]],
    T0 = sqrt(unlist(VarCorr(mod))),
    S =  sigma(mod)
  )
}
```


```{r eval = FALSE}
params <- expand.grid(
  rep = 1:10,
  beta = seq(0, 1, .25),
  n_trials = c(10, 20),
  n_subj = c(10, 20),
  tau_0 = seq(0, 1.5, .25),
  sigma = seq(0.5, 1.5, .25)
) %>%
  select(-rep)
y <- purrr::pmap_df(params, x)

save(y, file = "y")
```


```{r}
load("y")
```


```{r}
dat <- y %>%
  mutate(
    calc_avg_r = std_alpha2average_r(std.alpha, n_trials),
    prop_var = tau_0^2 / (tau_0^2 + sigma^2),
    calc_tau_0 = average_r2tau_0(calc_avg_r, sigma)) %>%
  group_by(n_subj, n_trials, beta, tau_0, sigma) %>%
  summarise(calc_tau_0 = mean(calc_tau_0, na.rm = T),
            std.alpha = mean(std.alpha, na.rm = T), 
            avg_r = mean(average_r, na.rm = T),
            prop_var = mean(prop_var, na.rm = T),
            calc_avg_r = mean(calc_avg_r, na.rm = T),
            T0 = mean(T0, na.rm = T),
            S = mean(S, na.rm = T),
            B = mean(B, na.rm = T),
            .groups = "drop")
```

## Parameter tau_0 versus estimate T0
```{r}
ggplot(dat, aes(tau_0, T0, color = as.factor(n_trials))) +
  geom_abline(slope = 1) + 
  geom_point(alpha = 0.5) + 
  facet_grid(beta~sigma, labeller = label_both)
```

## Parameter tau_0 versus estimate from std_alpha
```{r}
ggplot(dat, aes(tau_0, calc_tau_0, color = sigma)) +
  geom_abline(slope = 1) + 
  geom_point(alpha = .2) + 
  stat_summary(fun = "mean", geom = "point", size = 3) +
  facet_grid(~beta, labeller = label_both)
```

## Parameter sigma versus estimate S
```{r}
ggplot(dat, aes(sigma, S, color = tau_0)) +
  geom_point() + 
  geom_abline(slope = 1) + 
  facet_grid(n_trials~n_subj)
```

## Parameter beta versus estimate B
```{r}
ggplot(dat, aes(beta, B, color = tau_0)) +
  geom_abline(slope = 1) + 
  geom_point(alpha = .5) + 
  stat_summary(fun = "mean", geom = "point", size = 3) +
  facet_grid(sigma~tau_0, labeller = label_both)
``` 

## Estimate average_r versus calculated average_r
```{r}
ggplot(dat, aes(avg_r, calc_avg_r, color = sigma)) +
  geom_point() + 
  geom_abline(slope = 1) + 
  facet_grid(beta~sigma, labeller = label_both)
```

## Parameter proportion variance versus calculated average_r
```{r}
ggplot(dat, aes(prop_var, calc_avg_r, color = beta)) +
  geom_point() + 
  geom_abline(slope = 1) + 
  facet_grid(beta~sigma, labeller = label_both)
```
